Opening up the court (surface) in tennis grand slams

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

---
title: "Opening up the court (surface) in tennis grand slams"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{hyperref}
abstract: "The Australian Open, French Open, Wimbledon, and US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.  It is often thought that some players achieve better results at some grand slams than others and that difference in results may be attributed, at least partially, to the difference in the surfaces.  For example three top players of all time, Rafael Nadal, Roger Federer, and Serena Williams have achieved their best results on clay, grass, and hard courts, respectively.  In this paper, we examine the effect of surface type at grand slams from 2013-2017 for top players, adjusting for confounders such as league, competitor strength, and player attributes.  We answer the question of court effect on player results at first a high level using a ``complete'' data with few features and end using partial data with a variety of tennis-specific features.  Our primary model is a hierarchical model with fixed and random effects accounting for both player and tournament features.  We show that differences in surface type become more apparent with more detailed data and that more study should be conducted in this area."
bibliography: master.bib
---

@@

There has been little work on quantifying the high-level effect of court surface in tennis (issue 1). Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].   @knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass comparead to other courts. Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].

@@

The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).  There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two leagues, and seven rounds.  Each match has `r ncol(gs)` attributes, many of which are redundant.  We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.  Additionally, we take into account the number of sets in a match, the court surface, and round of the tournament.  A subset of the data is shown in Table \ref{tab:data}.

@@

We first determine whether the distribution of points comes from a normal distribution.  We plot the distribution of points in Figure \ref{fig:distr-eda} (left), and the distribution does seem approximately normal. This distribution is similar across tournament, with Wim. differing slightly from the other grand slams. As expected, there are more points earned in the WTA than the ATP (Fig. \ref{fig:distr-eda} (right)) due to the differing number of sets played. Also unsurprisingly, the winners of the match tended to earn more points than the losers.

@@

Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments. If there really is a home court advantage, the proportion of French wins each round would be higher at the FO than the AO, the USO, and Wim.. The same would be true for Australia and the US. However, in Figure \ref{fig:home-court} that this is not the case. After accounting for the number of players from each country, we do not find a home court advantage in the grand slams.

```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
                table(did_win, tournament)), 2)

with(filter(gs_players, ioc == "ESP", round == "R128"),
                table(did_win, tournament))

with(filter(gs_players, ioc == "ESP", round == "R128"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

@@

It is also commonly thought that Spaniards play better on clay. We are interested in whether Spaniards win matches at the French Open more than they win in other tournaments. It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in Figure \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments. However, this result is not significant (ENTER P VALUE). The ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse in ranking but still winning more than at the other tournaments.

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}

@@

ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))

@@

We find that the distribution of UEs and aces differ by tournament.  In Figure \ref{fig:ue-ace} we plot the total aces in a match vs. the total unforced errors in a match and color by the tournament. Specifically, matches in Wim. follow different ace and error patterns than the other tournaments. These matches tend to cluster in the upper left hand part of the graph, meaning that at Wim. players are making fewer unforced errors and scoring more aces. It is thought that the ball moves faster on grass courts and so the ace result is unsurprising, but making fewer errors is unexpected, especially because players seldom practice on grass.

@@

We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance. For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match. Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets). For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible covariates including country of origin (IOC), tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.

@@

```{r, fig.height = 3, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "none") + 
  xlab("") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5))
```

More broadly, the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated (see Table \ref{tab:corr}). This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other surfaces, but not among the hard courts and grass.

@@

```{r, fig.height = 2.5, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Aces") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  coord_flip() +
  theme(legend.position = "none")

winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Winners") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip() +
  theme(legend.position = "none")

nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Net Wins") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip() +
  theme(legend.position = "none")

ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Unforced Errors") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip() + 
  theme(legend.text=element_text(size=10))

grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))

@@

If we look at the random effects for Williams, Federer, and Nadal (see Fig. \ref{fig:mixed-mods-stats}), differences are much more noticeable than in the logistic regression model. For instance, we would expect Williams to have more aces at the AO and Wim. than at FO. We also see that Federer's performance at Wim. is consistently better than other tournaments in both aces and unforced errors. Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.

@@

We also fit individual linear models for Williams, Federer, and Nadal (subsetting the partial data to their matches only).  We estimate the percent of points won in a match using forward-backwards stepwise regression.  The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
$$
E[Y_{\% pts. won}|X] = \beta_0 + \beta_1 X_{opp. rank} + \beta_{2, AO}\mathcal{I}(X_{sur.} = AO)  + \beta_{2, Wim.}\mathcal{I}(X_{sur.} = Wim.) + \beta_{2, USO}\mathcal{I}(X_{sur.} = USO)  .
$$  
The full model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.  We do not display the full best fit models here, but they are available online.  We do report the effects ($\pm 2 \cdot$ Std. Err.) of significant variables and their interaction terms, adjusting for the other covariates.  Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically at AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.  For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.  For Williams, we also use forwards-backwards stepwise regression but begin at the low model since she has more missing data.

@@

For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.

@@

At the highest level, we find that the general point distribution seems to be the same for all court surfaces. There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g. that Spaniards perform better on clay). We do not find evidence for a home court advantage across all players in the grand slams. We also find that the FO seems to be the most different out of the four grand slams, but Wim. also displays some unique characteristics with regards to low amounts of UEs.

>>>>

---
title: "Opening up the court (surface) in tennis grand slams"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{hyperref}
abstract: "The Australian Open, The French Open, Wimbledon, and The US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.  It is often thought that some players achieve better results at some grand slams than others and that difference in results may be attributed, at least partially, to the difference in the surfaces.  For example three top players of all time, Rafael Nadal, Roger Federer, and Serena Williams have achieved their best results on clay, grass, and hard courts, respectively.  In this paper, we examine the effect of surface type at grand slams from 2013-2017 for top players, adjusting for confounders such as league, competitor strength, and player attributes.  We answer the question of court effect on player results at first a high level using ``complete'' data with few features and then using partial data with a variety of tennis-specific features.  Our primary model is a hierarchical model with fixed and random effects accounting for both player and tournament features.  We show that differences in surface type become more apparent with more detailed data and that more study should be conducted in this area."
bibliography: master.bib
---

@@

There has been little work on quantifying the high-level effect of court surface in tennis issue (1). Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001]. @knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass compared to other courts. Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].

@@

The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).  There are `r nrow(gs)` observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two leagues, and seven rounds.  Each match has `r ncol(gs)` attributes, many of which are redundant.  We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.  Additionally, we take into account the number of sets in a match, the court surface, and round of the tournament.  A subset of the data is shown in Table \ref{tab:data}.

@@

We first determine whether the distribution of points comes from a normal distribution.  We plot the distribution of points in Figure \ref{fig:distr-eda} (left), and the distribution does seem approximately normal. This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams. As expected, there are more points earned in the WTA than the ATP (Fig. \ref{fig:distr-eda} (right)) due to the differing number of sets played. Also unsurprisingly, the winners of the match tended to earn more points than the losers.

@@

Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments. If there really is a home court advantage, the proportion of French wins each round would be higher at the FO than the AO, the USO, and Wim.. The same would be true for Australia, the US, and Great Britain. However, in Figure \ref{fig:home-court} that this is not the case. After accounting for the number of players from each country, we do not find a home court advantage in the grand slams.

```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

chisq_r128 <- with(filter(gs_players, ioc == "ESP", round == "R128"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

chisq_r64 <- with(filter(gs_players, ioc == "ESP", round == "R64"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

chisq_r32 <- with(filter(gs_players, ioc == "ESP", round == "R32"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

chisq_r16 <- with(filter(gs_players, ioc == "ESP", round == "R16"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

@@

It is also commonly thought that Spaniards play better on clay. We are interested in whether Spaniards win matches at the French Open more than they win in other tournaments. It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in Figure \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments. However, this result is not significant at any round (p-value = `r round(chisq_r128$p.value, 2)`, `r round(chisq_r64$p.value, 2)`, `r round(chisq_r32$p.value, 2)`, `r round(chisq_r16$p.value, 2)`). The ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are worse in ranking yet winning more often.

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}

@@

ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))

@@

We find that the distribution of UEs and aces differ by tournament.  In Figure \ref{fig:ue-ace} we plot the total aces in a match vs. the total unforced errors in a match and color by the tournament. Specifically, matches in Wimbledon follow different ace and error patterns than the other tournaments. These matches tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces. It is thought that the ball moves faster on grass courts and so the ace result is unsurprising, but making fewer errors is unexpected, especially because players seldom practice on grass.

@@

We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance. For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match. Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets). For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible covariates including country of origin (IOC), tournament that the match was played at, whether the match occurred in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.

@@

```{r, fig.height = 3.5, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "none") + 
  xlab("") +
  ylab("Estimate") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5),
             top = textGrob("`ind_no_ioc' model",
                            gp=gpar(fontsize=20, fontfamily="serif")))
```

More broadly, the effects for the Australian Open, US Open, and Wimbledon are all quite correlated (see Table \ref{tab:corr}). This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other surfaces, but not among the hard courts and grass.

@@

```{r, fig.height = 3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Aces") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  coord_flip() +
  theme(legend.position = "none")

winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Winners") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip()  # +
  # theme(legend.position = "none")

nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Net Wins") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip() # +
  # theme(legend.position = "none")

ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Unforced Errors") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  theme(axis.text.y = element_blank()) +
  coord_flip()# + 
  #theme(legend.text=element_text(size=10))

#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))

ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
          common.legend = TRUE, legend="bottom", widths = c(.4, .3, .3)) %>%
grid.arrange(top = textGrob("Player-level effects",
                            gp=gpar(fontsize=20, fontfamily="serif")))

@@

If we look at the random effects for Williams, Federer, and Nadal (see Fig. \ref{fig:mixed-mods-stats}), differences are much more noticeable than in the logistic regression model. For instance, we would expect Williams to have more aces at the AO and Wim. than at FO. We also see that Federer's performance at Wimbledon is consistently better than other tournaments in both aces and unforced errors. Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.

@@

We also fit individual linear models for Williams, Federer, and Nadal (subsetting the partial data to their matches only).  We estimate the percent of points won in a match using forward-backwards stepwise regression.  The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
$$
E[Y_{\% pts. won}|X] = \beta_0 + \beta_1 X_{opp. rank} + \beta_{2, AO}\mathcal{I}(X_{sur.} = AO)  + \beta_{2, Wim.}\mathcal{I}(X_{sur.} = Wim.) + \beta_{2, USO}\mathcal{I}(X_{sur.} = USO)  .
$$  
The full model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.  We do not display the full best fit models here, but they are available online.  We do report the effects ($\pm 2 \cdot$ Std. Err.) of significant variables and their interaction terms, adjusting for the other covariates.  Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specifically at AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.  For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.  For Williams, we also use forwards-backwards stepwise regression but begin at the low model since she has more missing data.

@@

For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in opponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.

@@

At the highest level, we find that the general point distribution seems to be the same for all court surfaces. There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g. that Spaniards perform better on clay). We do not find evidence for a home court advantage across all players in the grand slams. We also find that the FO seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.
